---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r converttotime, echo = FALSE}
convert<- function(x){
        x = toString(x)
        while(nchar(x)<4) x <- paste0("0",x)
        chars <- strsplit(x,"")[[1]]
        x <- paste0(chars[1],chars[2],":",chars[3],chars[4])
}
```
Load the library and format the days and times for future graphs.
```{r load} 
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lubridate))
if(file.exists("activity.zip")) {
        if(!file.exists("activity.csv")) {
                unzip("activity.zip")
        }
        activities <- as_tibble(read.csv("activity.csv")) %>% rowwise() %>% mutate (tod = convert(interval)) %>% mutate( day = lubridate::wday(date, label = TRUE)) %>% mutate(weekend = (day == "Sat" || day == "Sun") )
        summary(activities$steps)
}
```
Code that formats the date
```{r converttotime, show = TRUE, eval = FALSE}
```
The data has a lot of 0's and N/As, I wanted to check if the NA's were by day or within the day.  If a whole day has NAs then it will be 288, if its all data then there will be 0 NAs.  This code pulls out all the other days.  Since 0 rows all NAs are grouped by day.
```{r}
activities <- mutate(activities, missing = is.na(steps))
as_tibble(with(activities, table(date, missing))) %>% filter( n != 288 & n != 0)
```
## What is mean total number of steps taken per day?
As you look at the summary, you can see there are a lot of zero's.
Zero is a valid step count for a 5 minute window, but unlikely for a day.
```{r} 
daily <- activities %>% ungroup() %>% group_by(date) %>% summarise(daily = sum(steps, na.rm = TRUE)) %>% filter(daily!= 0)
qplot(daily$daily, bins = 30)
step_mean <- mean(daily$daily, na.rm = TRUE)
step_median <- median(daily$daily, na.rm = TRUE)
```
The mean daily step total is: `r as.integer(step_mean)`  
The median daily step total is: `r as.integer(step_median)`  

## What is the average daily activity pattern?

```{r pattern, fig.width = 9}
segment <- activities %>% ungroup() %>% group_by(tod) %>% summarise(steps = sum(steps, na.rm = TRUE))
#with(segment, plot(factor(tod), steps, type="l" ))
ggplot(data = segment, aes(x = factor(tod), y = steps, group = 1))+geom_line() +theme(axis.text.x=element_text(color=rep(c("black", rep("transparent", each =11)), 24)))
peak_interval = segment %>% filter(steps == max(steps)) %>% select(tod)
```
Peak interval is at `r peak_interval`

## Imputing missing values
```{r nas}
numberNAs <- sum(is.na(activities$steps))
```
Data has `r numberNAs`.
Imputed data plan is to use the answer for that day of the week in that time interval.
```{r imputed}
daily_means <- activities %>% ungroup() %>% group_by(interval, day) %>% summarize(mean_steps = mean(steps, na.rm = TRUE))
imputed_activities = left_join(activities, daily_means, by = c("day", "interval")) %>% mutate( steps = ifelse(is.na(steps), mean_steps, steps))
imputed_daily <- imputed_activities %>% ungroup() %>% group_by(date) %>% summarise(daily = sum(steps, na.rm = TRUE)) %>% filter(daily!= 0)
imputed_step_mean <- mean(imputed_daily$daily, na.rm = TRUE)
imputed_step_median <- median(imputed_daily$daily, na.rm = TRUE)
```
Mean changed from `r as.integer(step_mean)` to `r as.integer(imputed_step_mean)`  
Median changed from `r as.integer(step_median)` to `r as.integer(imputed_step_median)`  

## Are there differences in activity patterns between weekdays and weekends?
```{r dailyplots, fig.width = 9}
segment <- activities %>% ungroup() %>% group_by(tod, weekend) %>% summarise(steps = sum(steps, na.rm = TRUE))

ggplot(data = segment, aes(x = factor(tod), y = steps, group = weekend))+geom_line( aes(color = weekend)) +theme(axis.text.x=element_text(color=rep(c("black", rep("transparent", each =23)), 12)))
```
Clearly there is a lot more activity during the week than there is on the weekend.
Just for fun lets look how much variety there is in the individual days.
```{r days, fig.width =11}
segment <- activities %>% ungroup() %>% group_by(tod, day) %>% summarise(steps = sum(steps, na.rm = TRUE))

ggplot(data = segment, aes(x = factor(tod), y = steps, group = day))+geom_line(aes(color = day)) +theme(axis.text.x=element_text(color=rep(c("black", rep("transparent", each =23)), 12)))
```
Thursday and Friday look more like a weekend in the AM than  Monday, Tuesday, Wednesday.  
